On the Limitations of Weighted Model Counting

### 2020-03-13 paper.tex
Moving on to the proof of the lower bound of extensions of BAs

\section{Introduction}

@@

\begin{theorem} \label{thm:extension} % TODO: cite the fact about atoms
  Let $\mathbf{B}$ be a finite Boolean algebra freely generated by some set of
  `literals' $L$, and let $m : \mathbf{B} \to \mathbb{R}_{\ge 0}$ be an
  arbitrary measure. We know that $\mathbf{B}$ has $n = 2^{|L|}$ atoms. Let
  $(a_i)_{i=1}^n$ denote those atoms in some arbitrary order. Let $L' = L \cup
  \{ \phi_i \mid i \in [n] \}$ be the set $L$ extended with $n$ new literals.
  Let $\mathbf{B'}$ be the unique Boolean algebra with
  \[
    \{ a_i \land \phi_i \mid i \in [n] \} \cup \{ a_i \land \neg \phi_i \mid i
    \in [n] \}
  \]
  as its set of atoms. Let $\iota : \mathbf{B} \to \mathbf{B'}$ be the inclusion
  homomorphism (i.e., $\iota(a) = a$ for all $a \in \mathbf{B}$). Let $w : L'
  \to \mathbb{R}_{> 0}$ be defined by
  \[
    w(l) = \begin{cases}
      \frac{m(a_i)}{2} & \text{if } l = \phi_i \text{ or } l = \neg\phi_i \text{
        for some } i \in [n] \\
      1 & \text{otherwise}
    \end{cases}
  \]
  for all $l \in L'$, and note that this defines a WMC measure $m' : \mathbf{B'}
  \to \mathbb{R}_{\ge 0}$. Then
  \[
    m(a) = (m' \circ \iota)(a)
  \]
  for all $a \in \mathbf{B}$.
\end{theorem}

@@

\begin{proof}
  % TODO
\end{proof}

% TODO: proof that n is the smallest number (in the worst case, without
% independence assumptions)
\begin{proposition}
  In the worst case (i.e., with no independence assumptions), adding $2^{|L|}$
  literals is the best we can do.
%  In \cref{thm:extension}, if $L'$ has less than $n$ additional elements, then
%  there exists a measure $$
\end{proposition}

>>>>

\section{Introduction}
% TODO: use \colon instead of :

@@

\begin{theorem} \label{thm:extension} % TODO: cite the fact about atoms
  Let $\mathbf{B}$ be a finite Boolean algebra freely generated by some set of
  `literals' $L$, and let $m : \mathbf{B} \to \mathbb{R}_{\ge 0}$ be an
  arbitrary measure. We know that $\mathbf{B}$ has $n = 2^{|L|}$ atoms. Let
  $(a_i)_{i=1}^n$ denote those atoms in some arbitrary order. Let $L' = L \cup
  \{ \phi_i \mid i \in [n] \} \cup \{ \neg \phi_i \mid i \in [n] \}$ be the set
  $L$ extended with $2n$ new literals. Let $\mathbf{B'}$ be the unique Boolean
  algebra with
  \[
    \{ \phi_i \land a_i \mid i \in [n] \} \cup \{ \neg \phi_i \land a_i \mid i
    \in [n] \}
  \]
  as its set of atoms. Let $\iota : \mathbf{B} \to \mathbf{B'}$ be the inclusion
  homomorphism (i.e., $\iota(a) = a$ for all $a \in \mathbf{B}$). Let $w : L'
  \to \mathbb{R}_{> 0}$ be defined by
  \[
    w(l) = \begin{cases}
      \frac{m(a_i)}{2} & \text{if } l = \phi_i \text{ or } l = \neg\phi_i \text{
        for some } i \in [n] \\
      1 & \text{otherwise}
    \end{cases}
  \]
  for all $l \in L'$, and note that this defines a WMC measure $m' : \mathbf{B'}
  \to \mathbb{R}_{\ge 0}$. Then
  \[
    m(a) = (m' \circ \iota)(a)
  \]
  for all $a \in \mathbf{B}$.
\end{theorem}

@@

% TODO: make J depend on i
\begin{proof} % TODO: find a reference for this first claim
  Since $\mathbf{B}$ is freely generated by $L$, each atom $a_i \in \mathbf{B}$
  is an infimum of elements in $L$, i.e.,
  \[
    a_i = \bigwedge_{j \in J} a_{i,j}
  \]
  for some $\{ a_{i,j} \}_{j \in J} \subset L$. Moreover, each atom $b \in
  \mathbf{B'}$ can be represented as either
  \[
    b = \phi_i \land a_i \quad \text{or} \quad b = \neg\phi_i \land a_i
  \]
  for some atom $a_i \in \mathbf{B}$, also making it an infimum over a subset of
  $L'$. Then, for any $b \in \mathbf{B}$,
  \[
    (m' \circ \iota)(b) = \sum_{\substack{\text{atoms } a_i \in \mathbf{B}:\\
        \phi_i \land a_i \le \iota(b)}} (w(\phi_i) + w(\neg\phi_i)) \prod_{j \in
    J} w(a_{i,j}),
  \]
  recognising that, for any $\iota(b)$, any atom $a_i \in \mathbf{B}$ satisfies
  \[
    \phi_i \land a_i \le \iota(b)
  \]
  if and only if it satisfies
  \[
    \neg\phi_i \land a_i \le \iota(b).
  \]
  Then, according to the definition of $w$,
  \[
    (m' \circ \iota)(b) = \sum_{\substack{\text{atoms } a_i \in \mathbf{B}:\\
        \phi_i \land a_i \le \iota(b)}} (w(\phi_i) + w(\neg\phi_i)) =
    \sum_{\substack{\text{atoms } a_i \in \mathbf{B}:\\ \phi_i \land a_i \le
        \iota(b)}} m(a_i) = m(b),
  \]
  provided that
  \[
    \phi_i \land a_i \le \iota(b) \quad \text{if and only if} \quad a_i \le b,
  \]
  but this is equivalent to
  \[
    \phi_i \land a_i = \phi_i \land a_i \land b \quad \text{if and only if}
    \quad a_i = a_i \land b
  \]
  which is true because $\phi_i \not\in L$.
\end{proof}

Now we can show that the construction in \cref{thm:extension} is smallest
possible.

% TODO: proof that n is the smallest number (in the worst case, without
% independence assumptions)
\begin{proposition}
  Let $\mathbf{B}$ and $\mathbf{B'}$ be Boolean algebras, and $\iota :
  \mathbf{B} \to \mathbf{B'}$ be the inclusion map such that $\mathbf{B}$ is
  freely generated by $L$, all atoms of $\mathbf{B'}$ can be expressed as
  meets of elements of $L'$, and the following subset relations are satisfied:
  \[
    \begin{tikzcd}
      \mathbf{B} \arrow{r}{\iota} & \mathbf{B'} \\
      L \arrow[Subset]{u}{} \arrow[Subset]{r}{} & L' \arrow[Subset]{u}{}
    \end{tikzcd}
  \]
  If, for any measure $m : \mathbf{B} \to \mathbb{R}_{\ge 0}$, one can
  construct a weight function $w : L' \to \mathbb{R}_{>0}$ such that the WMC
  measure $\WMC : \mathbf{B'} \to \mathbb{R}_{\ge 0}$ with respect to $w$
  satisfies
  \[
    m = \WMC \circ \iota,
  \]
  then $|L' \setminus L| \ge 2^{|L|+1}$.
\end{proposition}
\begin{proof}
  % 1. An atom in B' must have more than just elements of L.
  Let $a$ be an atom in $\mathbf{B}$, and let $b$ be an atom in $\mathbf{B'}$
  such that $b \le a$. First, let us notice that as long as $|L| \ge
  4$\footnote{Note that $|L|$ has to be an even number.}, $b \ne a$. Indeed, let
  $p, r, \neg p, \neg r \in L$. Then
  \begin{align*}
    (\WMC \circ \iota)(p \land r) &= w(p)w(r), \\
    (\WMC \circ \iota)(p \land \neg r) &= w(p)w(\neg r), \\
    (\WMC \circ \iota)(\neg p \land r) &= w(\neg p)w(r), \\
    (\WMC \circ \iota)(\neg p \land \neg r) &= w(\neg p)w(\neg r), \\
  \end{align*}
  But then we have that
  \[
    \frac{m(p \land r)}{m(\neg p \land r)} = \frac{w(p)}{w(\neg p)} =
    \frac{m(p \land \neg r)}{m(\neg p \land \neg r)}.
  \]
  This places a condition on $m$, contradicting the assumption that the
  construction works for an arbitrary $m$. Hence $b < a$.

  % 2. If we add phi, then we must also add -phi.
  % 3. Extension to multiple literals: we must have all (2^n) combinations of
  % added literals).
  % 4. Profit
\end{proof}
